Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-17 _tex/paper2.tex
Built site for gh-pages

\caption{\label{fig-example}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B1}.}

@@

Researchers can revisit the scenario in Section~\ref{sec-appendix-B1}
using these insights. By applying the BNF to the SCM
\ref{fig-scm_example1a} and DAG \ref{fig-mdag_example1a} structures,
they can represent the joint probability of the observed variables as
\(P(Y, T, X) =\) \(P(Y \:|\:T, X) P(T \:|\:X) P(X)\). From this
expression, researchers can infer that the outcome \(Y\) is not
unconditionally independent of the teaching method \(T\)
\((Y \:\bot\:T)\). This dependency arises from two key structures: a
direct causal path from the teaching method \(T\) to the outcome \(Y\),
represented by the two-connected-nodes structure \(T \rightarrow Y\)
(black path), and a confounding non-causal path from the teaching method
\(T\) to the outcome \(Y\) through the socio-economic status of the
school \(X\), represented by the fork structure
\(T \leftarrow X \rightarrow Y\) (gray path).

These probabilistic insights also indicate that narrowing the focus to a
subset of the population defined by the socio-economic status of the
school \(X\), achieved by conditioning on \(X\) and depicted by a square
in DAG \ref{fig-mdag_example1a}, effectively blocks all confounding
non-causal path introduced by the fork structure. Nevertheless, since
\(Y\) remains not d-separated from \(T\) even after conditioning
\((Y \:\not\bot\:T \:|\:X)\), this lack of d-separation can only be
attributed to the direct causal effect \(T \rightarrow Y\).

@@

\subcaption{\label{fig-scm_example1a}SCM}

@@

\subcaption{\label{fig-mdag_example1a}Conditioned DAG}

@@

\caption{\label{fig-example}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B1}.}

@@

With these new insights, researchers revisiting the scenario in
Section~\ref{sec-appendix-B22} can infer that the socio-economic status
of the school, \(X\), satisfies the backdoor criterion, assuming the
causal structure depicted by the SCM \ref{fig-scm_example1a} and DAG
\ref{fig-mdag_example1a} is correct. This means that \(X\) serves as a
sufficient adjustment set, as it effectively blocks all confounding
non-causal paths introduced by the fork structure. In this context,
researchers can apply the \emph{backdoor adjustment} to identify the ACE
of \(T\) on \(Y\). They achieve this by first identifying the CACE of
\(T\) on \(Y\) by conditioning on \(X\), and then marginalizing this
effect over \(X\) to obtain the ACE, as expressed in
Equation~\ref{eq-mACE2} (see Section~\ref{sec-appendix-B}). Notably, for
the purpose of identification, the conditioned DAG
\ref{fig-mdag_example1a} is equivalent to the manipulated DAG
\ref{fig-mdag_example1b}, because \(X\) satisfies the backdoor
criterion.

@@

\subcaption{\label{fig-scm_example1b}SCM}

@@

\subcaption{\label{fig-mdag_example1b}Manipulated DAG}

@@

\caption{\label{fig-example}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B22}.}

>>>>

\caption{\label{fig-example1}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B1}.}

@@

Researchers can revisit the scenario in Section~\ref{sec-appendix-B1}
using these insights. By applying the BNF to the SCM
\ref{fig-scm_example2} and DAG \ref{fig-mdag_example2} structures, they
can represent the joint probability of the observed variables as
\(P(Y, T, X) =\) \(P(Y \:|\:T, X) P(T \:|\:X) P(X)\). From this
expression, researchers can infer that the outcome \(Y\) is not
unconditionally independent of the teaching method \(T\)
\((Y \:\bot\:T)\). This dependency arises from two key structures: a
direct causal path from the teaching method \(T\) to the outcome \(Y\),
represented by the two-connected-nodes structure \(T \rightarrow Y\)
(black path), and a confounding non-causal path from the teaching method
\(T\) to the outcome \(Y\) through the socio-economic status of the
school \(X\), represented by the fork structure
\(T \leftarrow X \rightarrow Y\) (gray path).

These probabilistic insights also indicate that narrowing the focus to a
subset of the population defined by the socio-economic status of the
school \(X\), achieved by conditioning on \(X\) and depicted by a square
in DAG \ref{fig-mdag_example2}, effectively blocks all confounding
non-causal path introduced by the fork structure. Nevertheless, since
\(Y\) remains not d-separated from \(T\) even after conditioning
\((Y \:\not\bot\:T \:|\:X)\), this lack of d-separation can only be
attributed to the direct causal effect \(T \rightarrow Y\).

@@

\subcaption{\label{fig-scm_example2}SCM}

@@

\subcaption{\label{fig-mdag_example2}Conditioned DAG}

@@

\caption{\label{fig-example2}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B1}.}

@@

With these new insights, researchers revisiting the scenario in
Section~\ref{sec-appendix-B22} can infer that the socio-economic status
of the school, \(X\), satisfies the backdoor criterion, assuming the
causal structure depicted by the SCM \ref{fig-scm_example2} and DAG
\ref{fig-mdag_example2} is correct. This means that \(X\) serves as a
sufficient adjustment set, as it effectively blocks all confounding
non-causal paths introduced by the fork structure. In this context,
researchers can apply the \emph{backdoor adjustment} to identify the ACE
of \(T\) on \(Y\). They achieve this by first identifying the CACE of
\(T\) on \(Y\) by conditioning on \(X\), and then marginalizing this
effect over \(X\) to obtain the ACE, as expressed in
Equation~\ref{eq-mACE2} (see Section~\ref{sec-appendix-B}). Notably, for
the purpose of identification, the conditioned DAG
\ref{fig-mdag_example2} is equivalent to the manipulated DAG
\ref{fig-mdag_example3}, because \(X\) satisfies the backdoor criterion.

@@

\subcaption{\label{fig-scm_example3}SCM}

@@

\subcaption{\label{fig-mdag_example3}Manipulated DAG}

@@

\caption{\label{fig-example3}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B22}.}
